# Traffic Accident Patterns: A Comparative Analysis of US and Türkiye Traffic Data
---

# 1 Project Overview and Scope

## 1.1 Background

Traffic accidents are one of the major public safety problems worldwide and lead to significant human and economic losses every year. Understanding accident patterns and identifying factors associated with traffic accidents can help improve traffic safety and support better decision-making. Comparing accident data from different countries may also provide useful insights into similarities and differences in traffic behaviors and risk factors.

## 1.2 Problem Statement

Traffic accidents are influenced by many factors such as weather conditions, time of day, road conditions, and traffic density. However, the relationships among these factors are often difficult to understand without data analysis. This project aims to investigate traffic accident patterns and examine the factors associated with accident severity.

## 1.3 Objectives

The objectives of this project are:

- To analyze traffic accident patterns using a large-scale traffic accident dataset from the United States
- To compare accident trends between the United States and Türkiye
- To identify temporal and environmental factors associated with traffic accidents
- To investigate factors affecting accident severity
- To provide data-driven insights for traffic safety and decision-making

## 1.4 Research Questions

This project aims to answer the following questions:

1. How do traffic accident patterns change over time?
2. Which environmental and temporal factors are associated with accidents?
3. What factors affect accident severity?
4. Are there similarities between traffic accident trends between the United States and Türkiye?

## 1.5 Scope of the Study

This study uses publicly available traffic accident datasets from the United States and Türkiye. The US accident dataset will be used for detailed exploratory analysis, while traffic statistics from Türkiye will be used to establish comparisons and identify similarities and differences in traffic accident patterns.

---

# 2 Data

## 2.1 Data Source

This project uses two publicly available traffic accident datasets: one from the United States and one from Türkiye. The purpose of using these datasets together is to analyze detailed accident patterns using the US dataset while establishing a comparative connection with Türkiye.

The US Accidents dataset was obtained from Kaggle and covers traffic accidents across 49 states of the United States between February 2016 and March 2023. The dataset contains approximately 7.7 million accident records collected from multiple traffic information sources such as transportation departments, law enforcement agencies, traffic cameras, and traffic sensors.

The Türkiye traffic accident data were obtained from the Turkish Statistical Institute (TÜİK). The Türkiye data consist of yearly traffic accident statistics downloaded separately and include province-level accident information and weekday-based accident summaries.

Data sources:

-   US Accidents Dataset: https://www.kaggle.com/datasets/sobhanmoosavi/us-accidents
-   TÜİK Traffic Accident Statistics: https://data.tuik.gov.tr/

```{r}
#| code-fold: true
#| code-summary: "Show code"

library(tidyverse)
library(readr)
library(readxl)
library(lubridate)
library(purrr)

us_accidents <- read_csv(
  "data/US_Accidents_March23.csv",
  n_max = 100000,
  show_col_types = FALSE
)

accident_files <- list.files(
  "data/turkey_raw",
  pattern = "^accidents_.*\\.xlsx$",
  full.names = TRUE
)

weekday_files <- list.files(
  "data/turkey_raw",
  pattern = "^weekday_accidents_.*\\.xlsx$",
  full.names = TRUE
)
```

## 2.2 General Information About Data

The US dataset contains detailed accident-level information where each row represents a single accident event. The dataset includes variables related to accident severity, weather conditions, geographical information, and temporal characteristics.

Some important variables used in this study are:

| Variable          | Description                          |
|-------------------|--------------------------------------|
| Severity          | Severity level of the accident       |
| Start_Time        | Date and time of accident occurrence |
| Weather_Condition | Weather condition during accident    |
| Temperature(F)    | Temperature at the time of accident  |
| Visibility(mi)    | Visibility level                     |
| City              | Accident city                        |
| State             | Accident state                       |

The Türkiye dataset is more aggregated and focuses on yearly traffic statistics. It is mainly used to establish a comparison with US traffic patterns.

```{r}
#| code-fold: true
#| code-summary: "Show code"

dim(us_accidents)

print(accident_files)

print(weekday_files)
```

The output above shows the dimensions of the US dataset and confirms that all yearly Türkiye datasets were successfully imported into the project.

## 2.3 Reason of Choice

Traffic accidents are one of the most important public safety issues worldwide because they result in injuries, fatalities, and economic losses.

The US dataset was selected because it provides highly detailed accident information and supports analyses involving accident severity, weather conditions, and temporal patterns.

The Türkiye dataset was selected to establish a country-specific context and compare general accident characteristics between the two countries.

Using both datasets together improves the practical relevance of the study while providing richer analytical opportunities.

## 2.4 Preprocessing

Before starting the analysis, both datasets were prepared in R. Since the US dataset contains millions of records, a sample of 100,000 observations was used during the initial analysis stage to reduce computational burden.

The preprocessing steps applied are:

-   Importing datasets into R
-   Combining yearly Türkiye files
-   Selecting relevant variables
-   Creating time-related variables
-   Handling missing values
-   Preparing datasets for exploratory analysis

```{r}
#| code-fold: true
#| code-summary: "Show code"

turkey_accidents <- map_df(
  accident_files,
  read_excel
)

turkey_weekday <- map_df(
  weekday_files,
  read_excel
)

us_accidents_clean <- us_accidents %>%
  select(
    Severity,
    Start_Time,
    State,
    City,
    Weather_Condition,
    `Temperature(F)`,
    `Visibility(mi)`
  ) %>%
  mutate(
    Year = year(Start_Time),
    Month = month(Start_Time),
    Hour = hour(Start_Time),
    Weekday = wday(Start_Time, label = TRUE)
  ) %>%
  drop_na()

dim(us_accidents_clean)
dim(turkey_accidents)
dim(turkey_weekday)
```

The outputs above show the dimensions of the processed datasets that will be used in the exploratory and comparative analysis stages. 

# 3. Analysis


# Analysis

## 3.1 Exploratory Data Analysis

In this section, the US accident data are explored to understand general accident patterns in terms of time, location, severity, weather conditions, and environmental factors.

### 3.1.1 Yearly Accident Trend in the US

```{r}
#| code-fold: true
#| code-summary: "Show code"

us_yearly <- read_csv(
  "data/US_Accidents_March23.csv",
  col_select = Start_Time,
  show_col_types = FALSE
) %>%
  mutate(
    Year = year(Start_Time)
  ) %>%
  filter(
    Year < 2023
  ) %>%
  count(Year)

ggplot(
  us_yearly,
  aes(
    x = Year,
    y = n
  )
) +
geom_line(linewidth = 1) +
geom_point(size = 3) +
scale_x_continuous(
  breaks = unique(us_yearly$Year)
) +
labs(
  title = "Yearly Trend of Traffic Accidents in the US (2016–2022)",
  x = "Year",
  y = "Number of Accidents"
) +
theme_minimal()
```

The graph presents the yearly change in accident frequency across the US dataset. Records from 2023 were excluded because the dataset only contains accident data until March 2023. Including incomplete yearly data could create misleading trends.

### 3.1.1.1 Dataset Coverage by Year in the US

```{r}
#| code-fold: true
#| code-summary: "Show code"

us_coverage <- read_csv(
  "data/US_Accidents_March23.csv",
  col_select = c(Start_Time, State, City),
  show_col_types = FALSE
) %>%
  mutate(
    Year = year(Start_Time)
  ) %>%
  filter(
    Year < 2023
  ) %>%
  group_by(Year) %>%
  summarise(
    Number_of_States = n_distinct(State),
    Number_of_Cities = n_distinct(City)
  )

ggplot(
  us_coverage,
  aes(
    x = Year,
    y = Number_of_Cities
  )
) +
geom_line(linewidth = 1) +
geom_point(size = 3) +
scale_x_continuous(
  breaks = unique(us_coverage$Year)
) +
labs(
  title = "Number of Cities Covered in the US Dataset by Year",
  x = "Year",
  y = "Number of Cities"
) +
theme_minimal()
```

The increase in accident records between 2016 and 2022 cannot be explained solely by broader geographic coverage. The number of cities represented in the dataset increased by approximately 50% during this period, whereas the number of accident records increased by nearly four times. Therefore, the increase in accident records appears to be larger than the expansion in geographic coverage alone. This suggests that both dataset expansion and actual changes in accident patterns may contribute to the observed trend.

### 3.1.2 Traffic Accidents by Weekday in the US

```{r}
#| code-fold: true
#| code-summary: "Show code"

us_weekday <- us_accidents_clean %>%
  count(Weekday)

ggplot(us_weekday, aes(x = Weekday, y = n)) +
  geom_col() +
  labs(
    title = "Traffic Accidents by Weekday in the US",
    x = "Weekday",
    y = "Number of Accidents"
  ) +
  theme_minimal()
```

The graph shows that accident frequencies differ across weekdays. Higher accident frequencies during weekdays may be associated with commuting patterns and increased traffic volume, whereas lower frequencies during weekends may reflect reduced work-related travel activity.

### 3.1.3 Traffic Accidents by Hour of Day in the US

```{r}
#| code-fold: true
#| code-summary: "Show code"

us_hourly <- us_accidents_clean %>%
  count(Hour)

ggplot(us_hourly, aes(x = Hour, y = n)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  scale_x_continuous(breaks = 0:23) +
  labs(
    title = "Traffic Accidents by Hour of Day in the US",
    x = "Hour of Day",
    y = "Number of Accidents"
  ) +
  theme_minimal()
```

The hourly distribution indicates that accident frequencies are not evenly distributed throughout the day. Peaks during morning and evening hours may be associated with commuting periods when traffic density is typically higher.

### 3.1.4 Top Cities with Highest Accident Frequency in the US

```{r}
#| code-fold: true
#| code-summary: "Show code"

top_cities <- us_accidents_clean %>%
  count(City, sort = TRUE) %>%
  slice_head(n = 10)

ggplot(top_cities, aes(x = reorder(City, n), y = n)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top 10 Cities with Highest Accident Frequency in the US",
    x = "City",
    y = "Number of Accidents"
  ) +
  theme_minimal()
```

The results indicate that accident occurrences are concentrated in specific cities. Larger metropolitan areas with higher population density and traffic volumes are expected to exhibit greater accident frequencies.

### 3.1.5 Accident Severity Distribution in the US

```{r}
#| code-fold: true
#| code-summary: "Show code"

severity_distribution <- us_accidents_clean %>%
  count(Severity)

ggplot(severity_distribution, aes(x = factor(Severity), y = n)) +
  geom_col() +
  labs(
    title = "Accident Severity Distribution in the US",
    x = "Severity Level",
    y = "Number of Accidents"
  ) +
  theme_minimal()
```

The severity distribution demonstrates that accident records are concentrated in certain severity levels. Severe accidents occur less frequently, while moderate levels appear more common within the dataset.

### 3.1.6 Accident Severity by Weather Condition in the US

```{r}
#| code-fold: true
#| code-summary: "Show code"

top_weather <- us_accidents_clean %>%
  count(Weather_Condition, sort = TRUE) %>%
  slice_head(n = 8) %>%
  pull(Weather_Condition)

weather_severity <- us_accidents_clean %>%
  filter(Weather_Condition %in% top_weather) %>%
  count(Weather_Condition, Severity)

ggplot(
  weather_severity,
  aes(
    x = reorder(Weather_Condition, n),
    y = n,
    fill = factor(Severity)
  )
) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(
    title = "Accident Severity by Weather Condition in the US",
    x = "Weather Condition",
    y = "Number of Accidents",
    fill = "Severity"
  ) +
  theme_minimal()
```

The analysis suggests that weather conditions may influence accident severity. Different weather environments can affect road visibility and driving conditions, potentially leading to variations in accident outcomes.

### 3.1.7 Environmental Factors Associated with Accidents in the US

```{r}
#| code-fold: true
#| code-summary: "Show code"

environmental_factors <- us_accidents %>%
  summarise(
    Junction = sum(Junction, na.rm = TRUE),
    Traffic_Signal = sum(Traffic_Signal, na.rm = TRUE),
    Crossing = sum(Crossing, na.rm = TRUE),
    Railway = sum(Railway, na.rm = TRUE),
    Stop = sum(Stop, na.rm = TRUE),
    Give_Way = sum(Give_Way, na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "Factor",
    values_to = "Count"
  )

ggplot(environmental_factors, aes(x = reorder(Factor, Count), y = Count)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Environmental Factors Associated with Accidents in the US",
    x = "Environmental Factor",
    y = "Frequency"
  ) +
  theme_minimal()
```

The results indicate that accidents are associated with several environmental factors such as intersections, traffic signals, and crossings. These factors may contribute to accident occurrence due to increased traffic interactions and driving complexity.

### 3.1.8 Peak Hour Environmental Factors

```{r}
#| code-fold: true
#| code-summary: "Show code"

peak_hours <- us_accidents_clean %>%
  filter(Hour %in% c(7,8,9,16,17,18,19))

peak_environment <- peak_hours %>%
  summarise(
    Junction = sum(Junction, na.rm = TRUE),
    Traffic_Signal = sum(Traffic_Signal, na.rm = TRUE),
    Crossing = sum(Crossing, na.rm = TRUE),
    Railway = sum(Railway, na.rm = TRUE),
    Stop = sum(Stop, na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "Factor",
    values_to = "Count"
  )

ggplot(
  peak_environment,
  aes(
    x = reorder(Factor, Count),
    y = Count
  )
) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Environmental Factors During Peak Traffic Hours",
    x = "Factor",
    y = "Frequency"
  ) +
  theme_minimal()
```
The analysis indicates that accidents during peak traffic hours are more strongly associated with certain environmental factors such as intersections and traffic signals. Increased traffic interactions during commuting periods may increase the likelihood of conflicts between vehicles. Therefore, potential improvement strategies may include optimizing traffic signal timing, redesigning intersections, and implementing traffic flow management during peak hours.

## 3.2 Trend Analysis

This section will compare traffic accident patterns between the US and Türkiye.

## 3.3 Model Fitting

This section will include a simple model to investigate factors associated with accident severity.

## 3.4 Results

The results section will summarize the main findings from the exploratory analysis, trend comparison, and model fitting.